---
title: "Data Analysis With R"
author: "Mark Mayana"
date: "2024-12-10"
output:
  html_document:
    df_print: paged
---
Key Highlights
Orthodontic Data Analysis:

Objective: Analyze dental growth trajectories for children, grouped by gender.

Approach: Used lattice and nlme packages for visualizing grouped and individual growth patterns.

Outcome: Insights into gender-specific growth trends with well-structured line plots and grouped data visualizations.
Diamonds Dataset Exploration:

Objective: Examine relationships and distributions in the diamonds dataset.
Techniques:
Identified nominal and ratio variables for focused analysis.
Visualized distributions of price and depth, highlighting key trends (e.g., right-skewed price distribution).
Investigated the relationship between carat and price using scatter plots and regression analysis.
Key Insight: Strong positive correlation (0.92) between carat and price.
Telco Dataset Analysis:

Objective: Explore customer churn data for actionable insights.
Techniques:
Examined age-income correlation with scatter plots and annotations.
Visualized tenure distribution by gender using boxplots.
Computed the percentage of married individuals in Region 3.
Key Insights:
Weak correlation (0.33) between age and income.
Gender-wise tenure differences warrant further statistical testing.
Statistical Operations on Custom Dataset:

Ranked, sorted, and ordered a dataset, showcasing foundational data manipulation skills.
Skills and Tools Demonstrated
Libraries:
lattice, nlme, lme4, ggplot2, dplyr, grid, geepack, readxl, tinytex.
Statistical Techniques:
Correlation analysis, data transformations, and descriptive statistics.
Visualization:
Histograms, scatter plots, boxplots, and grouped line plots with annotations and captions.
Data Management:
Data cleaning, filtering, and organizing datasets for meaningful analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Required Libraries

```{r}
library(lme4)
library(nlme)
library(ggplot2)
library(dplyr)
library(lattice)
library(geepack)
library(readxl)
library(tinytex)
library(grid)  # For grid.text()

```

**Working with a Dataset**

```{r}
# Data input
data <- c(7, 4, 6, 8, 9, 1, 0, 3, 2, 5, 0)

# a) Rank the dataset
data_rank <- rank(data)

# b) Sort the data
data_sorted <- sort(data)

# c) Order the data (indices of sorted data)
data_order <- order(data)

# d) Create a data frame to view all results together
result <- data.frame(
  Original_Data = data,
  Rank = data_rank,
  Sorted_Data = data_sorted,
  Order = data_order
)

# Print the results
print(result)
```

**Analysis of the Diamonds Dataset in R library**

```{r}
data(diamonds)

# a) Identify two nominal variables and two ratio variables
nominal_vars <- c("cut", "color")  # Nominal variables
ratio_vars <- c("carat", "price") # Ratio variables

# Print the variables
print("Nominal variables: cut, color")
print("Ratio variables: carat, price")
```

b)  Plot a histogram of price values

```{r}
# Create the histogram
hist(
  diamonds$price,
  main = "Histogram of Diamond Prices",
  xlab = "Price (in US dollars)",
  col = "pink",
  border = "red",
  breaks = 50
)

# Add a comment in the bottom margin
mtext(
  "Comment: The price distribution is right-skewed, with most diamonds priced below $5000.",
  side = 1, # Bottom margin
  line = 4, # Distance from the axis
  col = "darkgreen" # Text color
)
```

c)  Determine the mean of the "depth" variable

```{r}
mean_depth <- mean(diamonds$depth)
print(paste("Mean depth:", mean_depth))

# Plot a histogram of diamond depths
hist(
  diamonds$depth,
  main = "Histogram of Diamond Depths",
  xlab = "Depth (%)",
  col = "yellow",
  border = "black",
  breaks = 30
)

mtext(
  "Comment:The depth distribution is bell-shaped, with most values between 57% and 65%.",
  side = 1, # Bottom margin
  line = 4, # Distance from the axis
  col = "darkgreen" # Text color
)
```

Scatter plot of carat vs price to check the relationship

```{r}
plot(
  diamonds$carat, diamonds$price,
  main = "Scatter Plot of Carat vs Price",
  xlab = "Carat",
  ylab = "Price (in US dollars)",
  col = "blue",
  pch = 16,
  cex = 0.5
)

# Add a regression line to the scatter plot
abline(lm(price ~ carat, data = diamonds), col = "black", lwd = 2)

# Calculate and print the correlation coefficient
correlation <- cor(diamonds$carat, diamonds$price)
print(paste("Correlation between carat and price:", correlation))

mtext(
  "The correlation coefficient (0.92) indicates a positive relationship between carat and price.",
  side = 1, # Bottom margin
  line = 4, # Distance from the axis
  col = "darkgreen" # Text color
)
```
**Telco Dataset Analysis**

```{r}
# Load customer churn dataset
file_path <- "C:/Users/USER/Downloads/School/Stastitical Computing sta1040/telco data.xlsx"
data <- read_excel(file_path, sheet = "telco")
```
Relationship between age and income

```{r}
# Calculate correlation coefficient
correlation_age_income <- cor(data$age, data$income, use = "complete.obs")
print(paste("Correlation between age and income:", round(correlation_age_income, 2)))

# Scatter plot for age vs income with annotation
ggplot(data, aes(x = age, y = income)) +
  geom_point(color = "blue", alpha = 0.6) +
  labs(
    title = paste("Scatter Plot of Age vs Income\nCorrelation Coefficient:", round(correlation_age_income, 2)),
    x = "Age",
    y = "Income",
    caption = "There is a weak correlation between age and income as indicated by the correlation coefficient.."
  ) +
  theme_minimal() +
  theme(
    plot.caption = element_text(color = "darkgreen", hjust = 1) # Customize caption
  )
```
Boxplot of tenure by gender

```{r}
ggplot(data, aes(x = factor(gender), y = tenure, fill = factor(gender))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Tenure by Gender",
    x = "Gender (0: Male, 1: Female)",
    y = "Tenure",
    caption = "The boxplot shows gender-wise differences in tenure, which can be analyzed further for significance."
  ) +
  scale_fill_manual(values = c("skyblue", "pink")) +
  theme_minimal() +
  theme(
    plot.caption = element_text(color = "darkgreen", hjust =0.1, size = 11)
  )
```
Percentage of married people residing in region 3

```{r}
# Filter data for Region 3
region3_data <- data %>% filter(region == 3)
total_region3 <- nrow(region3_data)
married_region3 <- sum(region3_data$marital == 1)

# Calculate and print percentage
married_percentage <- (married_region3 / total_region3) * 100
print(paste("Percentage of married people in Region 3:", round(married_percentage, 2), "%"))

```
**Orthodontic Data Analysis**
The Orthodontic dataset provides information on dental growth measurements.

```{r}
# Load Orthodont dataset
PRlong <- Orthodont

# Display the first few rows of the data
head(PRlong)
```
Open the data in a separate viewer

```{r}
View(PRlong)
```
Further analysis of the dataset

```{r}
# Show the structure of the data
str(PRlong)

# Display the dimensions of the data
dim(PRlong)

# Comment: The dataset contains orthodontic measurements, including age, gender, and distance values for analysis.

```
Gender column created to allow for simplified grouping and analysis based on sex.

```{r}
# Add a gender column
PRlong$gender <- ifelse(PRlong$Sex == "Male", "M", "F")
data.frame(PRlong$gender)
```

 Display the last 20 rows of selected columns

```{r}
tail(PRlong[, c("Subject", "age", "distance", "gender", "Sex")], 20)
```
Comment: The data is arranged by Subject for better visualization of individual records.

```{r}
# Arrange data by Subject
PRlong <- PRlong %>% arrange(Subject)

# View the arranged data
PRlong

```

Variables transformed for analysis, including scaling age and converting key fields to factors.

```{r}
# Perform transformations for further analysis
PRlong$Sage <- 2 * PRlong$age + 6 
PRlong$Sage.c.8 <- PRlong$age - 8

# Convert variables to factors for categorical analysis
PRlong$age <- as.factor(PRlong$age)
PRlong$Sgender <- as.factor(PRlong$gender)
PRlong$Subject <- as.factor(PRlong$Subject)
```
Createng a line plot of distance by age, grouped by gender

```{r}
# Line plot of distance by age, grouped by gender
xyplot(
  distance ~ age | gender,
  data = PRlong,
  type = "b",
  lwd = 2,
  pch = 19,
  cex = 1.2,
  groups = Subject
)

# Add caption for the xyplot
grid.text(
 "This plot shows individual dental growth trajectories by age, grouped by gender.",
  x = 0.5,  # Horizontal center
  y = 0.02, # Near bottom of the plot
  gp = gpar(col = "darkgreen", fontsize = 12) # Style the text
)

```
Grouped data plot for dental growth

```{r}
dental.grouped <- groupedData(distance ~ age | Subject, outer = ~ gender, data = PRlong)
plot(
  dental.grouped, 
  display = "subject", 
  outer = TRUE, 
  aspect = 1, 
  key = FALSE,
  xlab = "Age", 
  ylab = "Dental Growth (mm)", 
  pch = 19, 
  cex = 0.8,
  main = "Potthoff & Roy (1964) Orthodontic Measurements on Children"
)

# Add caption for the grouped data plot
grid.text(
  "This grouped data plot highlights dental growth trends across subjects and gender.",
  x = 0.5,  # Horizontal center
  y = 0.02, # Near bottom of the plot
  gp = gpar(col = "darkgreen", fontsize = 12, fontface = "italic") # Style the text
)

```
