---
title: "E-Commerce Sales Analysis"
author: "Mark Mayana"
date: "2024-12-19"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

**Introduction**
This analysis explores an e-commerce dataset to uncover insights related to revenue, sales transactions, customer behavior, product performance, and more. The dataset includes information on product pricing, categories, reviews, and discounts.

```{r setup}
# Load required libraries 
library(readxl)
library(readr)
library(dplyr)
library(tidyverse)
library(stringr)
library(ggplot2)
library(knitr)
library(kableExtra)

```
Load The Dataset

```{r}
data <- read_csv("C:/Users/USER/Downloads/datasets/amazon sales/amazon.csv")

# View structure and summary 
glimpse(data)
summary(data)
```
Data Cleaning and Handling Missing Values 

```{r}
# Check for missing values
colSums(is.na(data))

# Remove rows with missing values
data_clean <- data %>%
  drop_na()

# Check updated dataset
dim(data_clean)
colSums(is.na(data_clean))
```
Data Type Conversion of columns needed for analysis

```{r}
# Convert relevant columns to appropriate data types
data_clean <- data_clean %>%
  mutate(
    product_id = as.factor(product_id),
    category = as.factor(category),
    user_id = as.factor(user_id),
    review_id = as.factor(review_id),
    
    discounted_price = as.numeric(gsub("[^0-9.]", "", discounted_price)),
    actual_price = as.numeric(gsub("[^0-9.]", "", actual_price)),
    discount_percentage = as.numeric(gsub("[^0-9.]", "", discount_percentage)),
    rating = as.numeric(rating),
    
    rating_count = as.integer(rating_count),
    product_name = as.character(product_name),
    about_product = as.character(about_product),
    user_name = as.character(user_name),
    review_title = as.character(review_title),
    review_content = as.character(review_content),
    img_link = as.character(img_link),
    product_link = as.character(product_link),
    profit_margin = actual_price - discounted_price)

# View the cleaned structure
glimpse(data_clean)
```
**Exploratory Data Analysis**
Key Sales Metrics

```{r}
# Add a sales_amount column
data_clean <- data_clean %>%
  mutate(sales_amount = discounted_price)

# Calculate Key Metrics
# Total revenue
total_revenue <- sum(data_clean$sales_amount, na.rm = TRUE)

# Total number of transactions
total_transactions <- nrow(data_clean)

# Average and median sales values
avg_sales <- mean(data_clean$sales_amount, na.rm = TRUE)
median_sales <- median(data_clean$sales_amount, na.rm = TRUE)

# Display Metrics
key_metrics <- list(
  total_revenue = total_revenue,
  total_transactions = total_transactions,
  avg_sales = avg_sales,
  median_sales = median_sales
)
print(key_metrics)
```

Which Categories Generate the Most Revenue?

```{r}
category_revenue <- data_clean %>%
  group_by(category) %>%
  summarise(
    total_revenue = sum(sales_amount, na.rm = TRUE),
    avg_revenue = mean(sales_amount, na.rm = TRUE)
  ) %>%
  arrange(desc(total_revenue))

# View Category Results
print(category_revenue)
```

What are the top-selling and least-selling products
```{r}
# Adding short product names for better visual output
data_clean <- data_clean %>%
  mutate(short_product_name = str_extract(product_name, "^([^\\s]+\\s+){0,2}[^\\s]+"))

# Group and summarize products
top_products <- data_clean %>%
  group_by(short_product_name) %>%
  summarise(
    total_revenue = sum(sales_amount, na.rm = TRUE),
    total_quantity = sum(rating_count, na.rm = TRUE)
  ) %>%
  arrange(desc(total_revenue))

kable(top_products %>% head(10), caption = "Top 10 Products by Revenue")
kable(top_products %>% tail(10), caption = "Bottom 10 Products by Revenue")
```

Analyze total sales and revenue by product category
```{r}
# Analyze total sales and revenue by product category
category_performance <- data_clean %>%
  separate(category, into = c("main_category", "sub_category"), sep = "\\|", extra = "merge", fill = "right") %>%
  group_by(main_category) %>%
  summarise(
    total_revenue = sum(sales_amount, na.rm = TRUE),
    total_sales_volume = sum(rating_count, na.rm = TRUE)
  ) %>%
  arrange(desc(total_revenue))

# Print the results
print(category_performance)
```

**Customer Analysis**
Who are the top customers based on revenue or sales volume?
```{r}
# Analyze customer revenue
customer_analysis <- data_clean %>%
  mutate(first_name = str_extract(user_name, "^[^, ]+")) %>%
  group_by(first_name) %>%
  summarise(
    total_revenue = sum(sales_amount * rating_count, na.rm = TRUE),
    total_sales_volume = sum(rating_count, na.rm = TRUE)
  ) %>%
  arrange(desc(total_revenue))

kable(customer_analysis %>% head(10), caption = "Top 10 Customers by Revenue")
```
Are there repeat customers?
```{r}
# Add a new column with only the first name extracted from `user_name`
data_clean <- data_clean %>%
  mutate(first_name = str_extract(user_name, "^[^, ]+"))

repeat_customers <- data_clean %>%
  group_by(first_name) %>%
  summarise(transaction_count = n()) %>%
  filter(transaction_count > 1)

# View repeat customers
print(repeat_customers)
```


```{r}
# Segment customers based on purchase volume
customer_segments <- data_clean %>%
  group_by(user_id) %>%
  summarise(
    total_purchases = sum(rating_count, na.rm = TRUE),
    total_spent = sum(sales_amount, na.rm = TRUE)
  ) %>%
  mutate(
    customer_segment = case_when(
      total_purchases <= 5 ~ "Low Purchaser",
      total_purchases > 5 & total_purchases <= 15 ~ "Medium Purchaser",
      total_purchases > 15 ~ "High Purchaser"
    )
  ) %>%
  group_by(customer_segment) %>%
  summarise(
    avg_spent_per_customer = mean(total_spent, na.rm = TRUE),
    total_customers = n()
  )

# Print the results
print(customer_segments)
```
**Discount and Promotion Analysis**
Correlation Between Discounts and Sales Volume
```{r}
correlation <- cor(data_clean$discount_percentage, data_clean$rating_count, use = "complete.obs")
cat("Correlation between discount percentage and sales volume is:", correlation)
```
Discount Range Analysis
```{r}

discount_analysis <- data_clean %>%
  mutate(
    discount_range = case_when(
      discount_percentage >= 0 & discount_percentage < 20 ~ "0-20%",
      discount_percentage >= 20 & discount_percentage < 40 ~ "20-40%",
      discount_percentage >= 40 & discount_percentage < 60 ~ "40-60%",
      discount_percentage >= 60 ~ "60%+"
    )
  ) %>%
  group_by(discount_range) %>%
  summarise(
    total_sales_volume = sum(rating_count, na.rm = TRUE),
    total_revenue = sum(sales_amount, na.rm = TRUE),
    avg_rating = mean(rating, na.rm = TRUE)
  )

# Create a table with proper spacing between the title and content
discount_analysis %>%
  kable(caption = "Discount Range Analysis") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 12) %>%
  row_spec(0, extra_css = "padding-bottom: 15px;")  # Adds space below the header row
```

Grouping products by rating range
```{r}
rating_analysis <- data_clean %>%
  mutate(
    rating_range = case_when(
      rating >= 0 & rating < 3 ~ "Low (0-3)",
      rating >= 3 & rating < 4 ~ "Medium (3-4)",
      rating >= 4 ~ "High (4+)"
    )
  ) %>%
  group_by(rating_range) %>%
  summarise(
    total_sales_volume = sum(rating_count, na.rm = TRUE),
    total_revenue = sum(sales_amount, na.rm = TRUE)
  )

# Print the results
print(rating_analysis)

```



