---
title: 'STA1040: Group_Assignment_2'
author: "Group 1"
date: "2024-10-28"
output: html_document
---

```{r setup, include=FALSE}
# Load knitr package to set global chunk options
knitr::opts_chunk$set(warning = FALSE)
```

```{r}


# install.packages("tidyverse")

# load the packages
library(tidyverse)
library(lubridate)
library(gglpot2)
library(dplyr)
```

```{r}
# options(timeout = 300)  # Increase timeout to 300 seconds (5 minutes)
# download.file("https://dxl-datasets.s3.amazonaws.com/data/ppp_fl.csv", destfile = "ppp_fl.csv")
```

```{r}
# Load the downloaded CSV file
data <- read_csv("ppp_fl.csv")

```

```{r}

# viewing whole data in a new tab
view(data)
```

```{r}

# Dimension of dataframe
dim(data)
```

```{r}

# The first few rows
head(data)
```

```{r}
# The last few rows
tail(data)
```

```{r}
# To know all of our column names
colnames(data)
```

**Data cleaning**

```{r}
# Checking each variable type
glimpse(data)
```

As you can see from the above the data types of the variables: Date, Amount, AmountForgiven, BusinessAge, BusinessType need some modifications.

```{r}
# Changing the data type of the variable and setting a format
data$Date <- as.Date(data$Date, format = "%m/%d/%Y")

# Checking the change
class(data$Date)
```

```{r}
# Changing data type of the 2 variables to numeric
# But before converting them - we have to remove the $ and , sign
data$Amount <- gsub("[$,]", "", data$Amount)
data$AmountForgiven <- gsub("[$,]", "", data$AmountForgiven)

glimpse(data)
```

```{r}
# Changing the data types
data <- data %>% 
  mutate(Amount = as.numeric(data$Amount), 
         AmountForgiven = as.numeric(data$AmountForgiven))

# Renaming the columns to indicate the currency is in dollars
data <- data %>% 
  rename(Amount_in_Dollar = Amount, AmountForgiven_in_Dollar = AmountForgiven)

# Checking the change
class(data$Amount_in_Dollar)
class(data$AmountForgiven_in_Dollar)
```

For BusinessAge, and BusinessType we have to check whether we want them to be factor data types.

```{r}
# Extracting  unique values using unique(), and lapply() to appply the function to all the elements in the list 
unique_values <- lapply(list(BusinessAge = data$BusinessAge, 
                              BusinessType = data$BusinessType), unique)

# Print the unique values
print(unique_values)

```

```{r}
# Changing the data type of both variables in the above to factor
data <- data %>% 
  mutate(BusinessAge = as.factor(BusinessAge), 
         BusinessType = as.factor(BusinessType)) 

class(data$BusinessAge)
class(data$BusinessType)
```

```{r}
glimpse(data)
```

```{r}
# checking for missing Values
anyNA(data)
```

```{r}
#Checking for duplicated values
anyDuplicated(data)
```

```{r}
# Returns a logical vector indicating duplicated rows
duplicates <- duplicated(data)

# Show only duplicated rows
data[duplicates, ]
```

**Breaking down the Borrower Address Column**

This allows for analysis of geographic distribution, grouping by state or city, and identifying regional trends

```{r}

# Split BorrowerAddress into street, city, and a combined state/zip column
data <- data %>%
  separate(BorrowerAddress, into = c("BorrowerStreet", "BorrowerCity", "BorrowerStateZip"), sep = ",\\s*", extra = "merge", fill = "right") %>%
  separate(BorrowerStateZip, into = c("BorrowerState", "BorrowerZip_Code"), sep = "\\s", extra = "merge", fill = "right")

# Preview the resulting dataset with the new address columns
View(data)
```

```{r}
# Ensure ZIP codes are in the correct format (5 digits)
data <- data %>%
  mutate(BorrowerZip_Code = substr(BorrowerZip_Code, 1, 5))
```

```{r}
glimpse(data)
```

**Breaking down the Date Column**

This breakdown is important because having separate columns for year, month, and day enables us to analyze data at a more granular level, such as trends over time or seasonal effects.

```{r}

data <- data %>%
  mutate(Date = ymd(Date), # Converting the Date to proper date format (MM/DD/YYYY)
         Year = year(Date), # Extracting the year
         Month = month(Date), # Extracting the month
         Day = day(Date)) # Extracting the day

# View the resulting dataset
View(data)
```

**Breaking down the Lender Address Column**

By separating components like street, city, state, and zip code, you can analyze each aspect individually. For example, you can focus on city-wise trends, which might not be apparent when the addresses are combined.

```{r}
# Separate LenderAddress into multiple components and add them as new columns
data <- data %>%
  separate(LenderAddress, into = c("LenderStreet", "LenderCity", "LenderStateZip"), sep = ",\\s*", extra = "merge", fill = "right") %>%
  separate(LenderStateZip, into = c("LenderState", "LenderZipCode"), sep = "\\s", extra = "merge", fill = "right")

# Preview the resulting dataset with the new address columns
View(data)

```

```{r}
# Ensure ZIP codes are in the correct format (5 digits)
data <- data %>%
  mutate(LenderZipCode = substr(LenderZipCode, 1, 5))

# Preview the resulting dataset
View(data)
```

```{r}
# Save the cleaned dataset as a CSV file 
write.csv(data, "cleaned_dataset.csv", row.names = FALSE)
```

```{r}
# Import the cleaned dataset
cleaned_data <- read_csv("cleaned_dataset.csv")
```

```{r}
View(cleaned_data)
```

```{r}
glimpse(cleaned_data)
```

```{r}
# Changing the data type of both variables in the above to factor
cleaned_data <- cleaned_data %>% 
  mutate(BusinessAge = as.factor(BusinessAge), 
         BusinessType = as.factor(BusinessType)) 

class(cleaned_data$BusinessAge)
class(cleaned_data$BusinessType)
```


```{r}
glimpse(cleaned_data)
```

### Question 2: Return the total amount of money loaned by month from 2020 to 2021.

```{r}
# First let's add a new "Total_loan_in_dollar" column 
cleaned_data <- cleaned_data %>% 
 mutate(Total_loan_in_dollar = Amount_in_Dollar + AmountForgiven_in_Dollar)

 #cleaned_data$Total_loan_in_dollar <- cleaned_data$Amount_in_Dollar + cleaned_data$AmountForgiven_in_Dollar
 #view(cleaned_data)
```

```{r}
# Convert numeric months to full month names
cleaned_data <- cleaned_data %>%
  mutate(Month = month.name[Month])
```

```{r}
# Grouping by month and returning the total loan
grouped_data <- cleaned_data %>%
  filter(Year %in% c(2020, 2021)) %>%
  group_by(Month) %>% 
  summarise(Total_amount_of_money_loaned_from_2020_to_2021 = sum(Total_loan_in_dollar))

# View the grouped data
view(grouped_data)

```

```{r}
# Define the correct order for the months
month_order <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")

# Convert the Month column to a factor with the specified levels
grouped_data$Month <- factor(grouped_data$Month, levels = month_order)

# Order the data frame by Month
ordered_data <- grouped_data[order(grouped_data$Month), ]

view(ordered_data)
```


### Question 3: Return the name and total money loaned for Florida's top PPP lender in 2021 ("top" = most money loaned)

```{r}
# Filtering lenders in FL in 2021, summing the loan amounts
florida_top_lender <- cleaned_data %>% 
  filter(Year == 2021 & LenderState == "FL")  %>% 
  group_by(Lender) %>% 
  summarise(Total_loan_in_dollar = sum(Total_loan_in_dollar, na.rm = TRUE)) %>%
  arrange(desc(Total_loan_in_dollar))  # Ordering from highest to lowest

# View the sorted lender data
view(florida_top_lender)

# Get the lender with the maximum total loan amount
top_lender <- florida_top_lender %>%
  slice_max(Total_loan_in_dollar)  # Alternative to filter for max value
top_lender

``` 

### Question 4: what percent of loan money went to borrowers in the state of Miami?

```{r}
# Filter for Miami and calculate total loan amount for Miami
miami_total <- data %>%
  filter(BorrowerCity == "Miami") %>%
  summarise(TotalMiamiLoan = sum(Amount_in_Dollar, na.rm = TRUE))

# Calculate total loan amount for all borrowers
total_loan_amount <- data %>%
  summarise(TotalLoan = sum(Amount_in_Dollar, na.rm = TRUE))

# Calculate the percentage of loan money for Miami
miami_percentage <- (miami_total$TotalMiamiLoan / total_loan_amount$TotalLoan) * 100

# Display the result
miami_percentage
```
### Question 5: Bar plot showing average jobs by business age in certain groups

```{r}

# Categorize Business Age into the four specified categories
data <- data %>%
  mutate(BusinessAgeCategory = recode(BusinessAge,
                                      "Startup, Loan Funds will Open Business" = "Startup",
                                      "2 years old or less" = "0-2 years",
                                      "Existing or more than 2 years old" = "3+ years",
                                      .default = "Other/Unknown"
  ))

# Calculate the average jobs by business age category and drop the groups
agg_tbl <- data %>%
  group_by(BusinessAgeCategory) %>%
  summarise(Average_Jobs = mean(Jobs), .groups = 'drop')

# Reorder the BusinessAgeCategory factor based on Average_Jobs in descending order
agg_tbl <- agg_tbl %>%
  mutate(BusinessAgeCategory = fct_reorder(BusinessAgeCategory, Average_Jobs, .desc = TRUE))

# Plot the average jobs by business age category, now ordered from highest to lowest
ggplot(agg_tbl, aes(x = BusinessAgeCategory, y = Average_Jobs)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme_minimal() +
  labs(title = "Average Jobs by Business Age Category", x = "Business Age Category", y = "Average Jobs") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

### Question 6: Horizontal bar plot showing total amount borrowed in each zip code in the city of Coral Gables

```{r}
# Filter for Coral Gables and only 5-digit zip codes
coral_gables_data <- cleaned_data %>%
  filter(BorrowerCity == "CORAL GABLES" & nchar(as.character(BorrowerZip_Code)) == 5)

# Summarize the total amount borrowed by zipcode
zipcode_summary <- coral_gables_data %>%
  group_by(BorrowerZip_Code) %>%
  summarise(Total_Amount_Borrowed = sum(Amount_in_Dollar, na.rm = TRUE)) %>%
  arrange(desc(Total_Amount_Borrowed))

# Convert amount to millions for better readability
zipcode_summary$Total_Amount_Borrowed <- zipcode_summary$Total_Amount_Borrowed / 1e6

# Create the horizontal bar plot
ggplot(zipcode_summary, aes(x = reorder(BorrowerZip_Code, Total_Amount_Borrowed), y = Total_Amount_Borrowed)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +  # Flip coordinates for horizontal bars
  scale_y_continuous(labels = scales::dollar_format(scale = 1, prefix = "$", suffix = "M")) +  # Format y-axis labels
  labs(title = "Total Amount Borrowed in Coral Gables by Zip Code",
       x = "Zip Code",
       y = "Total Amount Borrowed (in Millions)") +
  theme_minimal()

```
